---
title: "Interactive and explorative visualization of ExpressionSet using ExpressionSetViewer"
output: 
  html_document:
  # rmarkdown::html_vignette
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
number_sections: true
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

### ExpressionSet

ExpressionSet is the S4 object storing high throughput omics data defined in Bioconductor. The core component
of ExpressionSet is an expression matrix, where the rows are features, such as genes, proteins, and columns are
samples. The values in the matrix represent the abundance or presence/absence of features. The meta-information about
features (rows) and samples (columns) are store in data.frames called feature data and phenotype data, respectively. 
More detailed instruction of ExpressionSet could be found here:
https://www.bioconductor.org/packages/release/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf

### ExpressionSet viewer

ExpressionSetViewer visualizes ExpressionSet in an interactive way to facilitate data exploration. The ExpressionSetViewer has separate back- and front-end. In the back-end, users need to prepare an ExpressionSet that contains all the necessary information for the downstream data interpretation. Some extra requirement on the headers of phenotype data or feature data are imposed so that the provided information can be clearly recognized by the front-end, at the same time, keep a minimum modification on the existing ExpressionSet object. The pure dependency on R/Bioconductor guarantees maximum flexibility in the statistical analysis in the back-end. Once the ExpressionSet is prepared, it can be visualized using the front-end, implemented by shiny and plotly. Both features and samples could be selected from (data) tables or graphs (scatter plot/heatmap). Different types of analyses, such as enrichment analysis (using Bioconductor package fgsea or fisher's exact test) and STRING network analysis, will be performed on-the-fly and the results are visualized simultaneously. When a subset of samples and a phenotype variable is selected, a significance test on means (t-test or ranked based test; when phenotype variable is quantitative) or test of independence (chi-square or fisher’s exact test; when phenotype data is categorical) will be performed to test the association between the phenotype of interest with the selected samples. Therefore, ExpressionSetViewer will greatly facilitate data exploration, many different hypotheses can be explored in a short time without the need for knowledge of R. 

## A quick start

Once the data for ExpressionSetViewer be done, the data could be visualized separately. This greatly 
facilitate the commutation between bioinformaticians and experiemental biologist. To open the 
ExpressionSetViewer with shiny:

```{r eval=FALSE}
library(ExpressionSetViewer)
path2data <- system.file("extdata", package = "ExpressionSetViewer")
ExpressionSetViewer(path2data)
```

Here the path2data is folder that containing the .RDS files of ExpressionSetViewer. 
All the .RDS files will be listed in the text box on the top right corner. Select the 
data you want to load there. 

A video introduction on how to use the viewer is here:

## New requirement on the ExpressionSet

The ExpressionSetViewer tries to make the changes on ExpressionSet object as small
as possible. In order to let the users quickly navigate the information they want to 
check, there are two general requirements. 
 - The expression matrix should have unique row names and column names. The row names
 of feature data should be the same as row names of expression matrix. The row names
 of phenotype data should be the same as column names of expression matrix. It is also 
 recommand to use meaningful names such as gene names, because the names will be displayed
 in the expression viewer. 
 - The columns of feature data and phenotype data should be in this three elements format:
   "Analysis|Subset|Variable"
   In this way, the variables could be selected with "three step selector", the first 
   elements is call analysis, for example "t-test", "PCA". The second element is subset, it 
   often indicates on which subset of samples/feature the analysis was played. For example, if in 
   an experiment with three groups, e.g. A, B and C, a t-test was used to compare the group
   A and B, the first two elements could be "ttest|A_vs_B". The last elements are the variables
   return by the analysis, for example the fold change of 
   
### The reserved headers
 - **Surv** in phenotype data, this 
 - **StringDB** in the feature data

<!--
## More detailed instruction

### General information

### Gene set information
  
### Survival data

### t-test

### PCA

### StringDB connection

## Analysis can be done inside ExpressionSetViewer
-->

## deployment

In addition, the resulted data could be easily shared using a shiny server. Otherwise, a standalone version of ExpressionSetViewer together with designated omics data could be easily created by integrating it with portable R, which can be shared with collaborators or submitted as supplementary data together with a manuscript. 

### Standalone version using docker

Docker is the easiest way to share the with docker user

### standalone version using R portable

To share the ExpressionSetViewer and data with windows user, you can install the r portable
https://sourceforge.net/projects/rportable/


### shiny server

ExpressionSetViewer could be easily deployed using shiny server. For example, the ui file

```{r deploy_shiny_ui,eval=FALSE}
library(shiny)
ui <- fluidPage(
  ExpressionSetViewer:::app_ui("app")
)
```

The server file

```{r deploy_shiny_server,eval=FALSE}
# dir need to be changed
dir <- "path/to/object"
server <- function(input, output, session) {
  callModule(ExpressionSetViewer:::app_module, id = "app", dir = reactive(dir))
}
```

### inside R

Assume you have prepared an object for ExpressionSetViewer, you can open it in R using

```{r deploy_inside_R,eval=FALSE}
library(ExpressionSetViewer)
ExpressionSetViewer("path/to/object")
```

The object could be loaded from the top-right corner. 

### Comparison of different deployment methods

## A simple example workflow

A simple pipeline:
```{r simple_pipeline}
library(ExpressionSetViewer)
packdir <- system.file("extdata", package = "ExpressionSetViewer")
# reading expression
expr <- read.delim(file.path(packdir, "expressionMatrix.tsv"), stringsAsFactors = FALSE)
colnames(expr) <- make.names(colnames(expr))
rownames(expr) <- make.names(rownames(expr))
# reading feature data
fd <- read.delim(file.path(packdir, "featureGeneral.tsv"), stringsAsFactors = FALSE)
# reading phenotype data
pd <- read.delim(file.path(packdir, "sampleGeneral.tsv"), stringsAsFactors = FALSE)

#  reading other datasets
drugData <- read.delim(file.path(packdir, "sampleDrug.tsv"))
# survival data
# this data is from cell line, the survival data are fake data to show how to use the survival data in ExpressionSetViewer
surv <- read.delim(file.path(packdir, "sampleSurv.tsv"))
# gene set information
genesets <- read.delim(file.path(packdir, "featureGS.tsv"))

# Define t-test to be done
tests <- rbind(
 c("Origin", "RE", "ME"),
 c("Origin", "RE", "LE"),
 c('TP53.Status', "MT", "WT")
 )

# prepare column for stringDB query
strid <- sapply(strsplit(fd$Protein.ID, ";|-"), "[", 1)

###
d <- prepEsetViewer(
  expr = expr, pData = pd, fData = fd, 
  PCA = TRUE, pca.fillNA = TRUE,
  t.test = tests, ttest.fillNA = FALSE, 
  gs = genesets, stringDB = strid, surv = surv)

# feature space - default x axis
attr(d, "fx") <- "ttest|RE_vs_ME|mean.diff"
# feature space - default y axis
attr(d, "fy") <- "ttest|RE_vs_ME|log.fdr"
# sample space - default x axis
attr(d, "sx") <- "PCA|All|PC1("
# sample space - default y axis
attr(d, "sy") <- "PCA|All|PC2("

# saveRDS(d, file = "dtest.RDS")
##  to open the viewer
# ExpressionSetViewer("./")
```

## sessionInfo
```{r session}
sessionInfo()
```
